---
title: "STA3050A - Assignment 1"
author: "Chesia Anyika"
date: "2024-05-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Question

The annual expenditure levels (in millions) to promote products and services for the financial services sector such as banks , insurance, investments, etc. from 2015 to 2022 are shown in the following table:

| Year            | 2015 | 2016 | 2017 | 2018 | 2019 | 2020 | 2021 | 2022 |
|-----------------|------|------|------|------|------|------|------|------|
| **Expenditure** | 5.5  | 7.2  | 8.0  | 9.6  | 10.2 | 11.0 | 12.5 | 14.0 |

Use exponential smoothing to obtain forecasted values by taking $\alpha = 0.5, 0.7$ and $0.9$ and calculate the forecast errors. Also plot the original and smoothed values.

# Execution

## Libraries

```{r}
#load the necessary libraries
library(tidyverse)
```

## Create Table in Long Format

First I input the given values into vectors, and created a table in **long format** using the `data.frame()` function.

```{r}
#create vectors
year <- c('2015','2016','2017','2018','2019','2020','2021','2022')
expenditure <- c(5.5, 7.2, 8.0, 9.6, 10.2 ,11.0, 12.5, 14.0)

#create table
data <- data.frame(year, expenditure)

#view table
data
```

## Case 1: $\alpha = 0.5$

### Step 1: Obtain Forecast Values Using Exponential Smoothing

The formula for obtaining filtered values using exponential smoothing is:

$$
y't = \alpha yt + (1- \alpha)y't - 1
$$

To implement this formula in r, I first defined my alpha parameter, then I initialised an empty column in the `data` table for forecasted values, and specified the value of the first forecast value as 5.5. This is because the first forecast value ( $y't$ ) is always equals to the first value of the variable of interest ( $yt$ ), represented mathematically as: $y'1 = y1$.

```{r}
#define alpha parameter
alpha <- 0.5

# Initialize the first forecast value
data$Forecast.5 <- NA
data$Forecast.5[1] <- data$expenditure[1]

#view the results
data
```

I then created a **for loop** that calculates the forecast values as per the formula given above, using the **value of interest in the expenditure column (** $yt$ **)**, and the **preceding value computed in the forecast column (** $y't$ **).** This yields a column of the computed forecast values with $\alpha = 0.5$.

```{r}
# Calculate the forecasted values using exponential smoothing
for (i in 2:nrow(data)) {
  data$Forecast.5[i] <- alpha * data$expenditure[i] + (1 - alpha) * data$Forecast.5[i-1]
}

# View the dataframe with forecasted values
data
```

### Step 2: Compute Forecast Error

Forecast error is computed as $yt - y't$. I computed this as follows:

```{r}
# Calculate the forecast error for each time period
data$Error.5 <- data$expenditure - data$Forecast.5

data
```

## Case 2: $\alpha = 0.7$

### Step 1: Obtain Forecast Values Using Exponential Smoothing

I defined my alpha parameter as 0.7, initialised a new empty column for the forecast values and specified the first value. Then, I used the same for loop of the exponential smoothing formula created in Step 2 to compute the subsequent values.

```{r}
#define alpha parameter
alpha <- 0.7

# Initialize the first forecast value
data$Forecast.7 <- NA
data$Forecast.7[1] <- data$expenditure[1]

#view the results
data
```

```{r}
# Calculate the forecasted values using exponential smoothing
for (i in 2:nrow(data)) {
  data$Forecast.7[i] <- alpha * data$expenditure[i] + (1 - alpha) * data$Forecast.7[i-1]
}

# View the dataframe with forecasted values
data
```

### Step 2: Compute Forecast Error

Forecast error is computed as $yt - y't$. I computed this as follows:

```{r}
# Calculate the forecast error for each time period
data$Error.7 <- data$expenditure - data$Forecast.7

data
```

## Case 3: $\alpha = 0.9$

### Step 1: Obtain Forecast Values Using Exponential Smoothing

I defined my alpha parameter as 0.9, initialised a new empty column for the forecast values and specified the first value. Then, I used the same for loop of the exponential smoothing formula created in Step 2 to compute the subsequent values.

```{r}
#define alpha parameter
alpha <- 0.9

# Initialize the first forecast value
data$Forecast.9 <- NA
data$Forecast.9[1] <- data$expenditure[1]

#view the results
data
```

```{r}
# Calculate the forecasted values using exponential smoothing
for (i in 2:nrow(data)) {
  data$Forecast.9[i] <- alpha * data$expenditure[i] + (1 - alpha) * data$Forecast.9[i-1]
}

# View the dataframe with forecasted values
data
```

### Step 2: Compute Forecast Error

Forecast error is computed as $yt - y't$. I computed this as follows:

```{r}
# Calculate the forecast error for each time period
data$Error.9 <- data$expenditure - data$Forecast.9

data
```

## Compare Accuracy of Forecast Values

I then used the forecast values to compute **Mean Absolute Error** and **Mean Squared Error**, to compare the accuracy of the different forecast values given the alpha values.

```{r}
# Compute the Mean Absolute Error (MAE)
MAE.5 <- mean(abs(data$Error.5), na.rm = TRUE)
MAE.7 <- mean(abs(data$Error.7), na.rm = TRUE)
MAE.9 <- mean(abs(data$Error.9), na.rm = TRUE)

# Compute the Mean Squared Error (MSE)
MSE.5 <- mean(data$Error.5^2, na.rm = TRUE)
MSE.7 <- mean(data$Error.7^2, na.rm = TRUE)
MSE.9 <- mean(data$Error.9^2, na.rm = TRUE)


# View MAE and MSE
cat('\n Mean Absolute Errors: ', '\n alpha = 5: ', MAE.5, '\n alpha = 7: ', MAE.7, '\n alpha = 9: ', MAE.9, '\n Mean Squared Errors: ', '\n alpha = 5: ', MSE.5, '\n alpha = 7: ', MSE.7, '\n alpha = 9: ', MSE.9)

```

**Interpretation**

Lower MAE and MSE values indicate a more accurate forecast value. For both MAE and MSE:

-   Case 3 with $\alpha = 0.9$ has the most accurate forecast values. (MAE = 0.1158, MSE = 0.0167)

-   Ccase 2 with $\alpha = 0.7$ has less accurate forecast values. (MAE = 0.4223, MSE = 0.2137)

-   Case 1 with $\alpha = 0.5$ has the least accurate forecast values. (MAE = 0.8952, MSE = 0.9497)

## Plot the Original and Smoothed Values

I plotted the original and forecast values as **line graphs**, with the original values being a **solid line** and the Forecast values being **dashed lines** of **different colours dependent on the alpha value.** I did this usign the `ggplot()` and `geom_line()` functions as shown below:

```{r}
# Plot original and forecasted values
ggplot(data, aes(x = year)) +
  geom_line(aes(y = expenditure, color = "Original", group = 1), linetype = "solid") +
  geom_line(aes(y = Forecast.5, color = "Forecasted (alpha = 0.5)", group = 1), linetype = "dashed") +
  geom_line(aes(y = Forecast.7, color = "Forecasted (alpha = 0.7)", group = 1), linetype = "dashed") +
  geom_line(aes(y = Forecast.9, color = "Forecasted (alpha = 0.9)", group = 1), linetype = "dashed") +
  labs(title = "Original and Forecasted Expenditure Values",
       y = "Expenditure (millions)",
       color = "Series") +
  scale_color_manual(values = c("Original" = "black",
                                 "Forecasted (alpha = 0.5)" = "red",
                                 "Forecasted (alpha = 0.7)" = "blue",
                                 "Forecasted (alpha = 0.9)" = "green")) +
  theme_minimal()

```

**Interpretation**

The above visualisation shows that:

-   Case 3 values forecast using $\alpha = 0.9$ exhibit the least difference form the original values, as evidenced by how close their respective lines are on the graph.

-   Case 2 values forecast using $\alpha = 0.7$ exhibit more of a difference from the original values.

-   Case 3 values forecast using $\alpha = 0.9$ exhibit the greatest difference from the original values, as evidenced by how the two respective lines are the furthest away from each other on the graph.

These observations are in line with the **observations made when comparing the accuracy of the models**. Thus, it can be concluded that the higher the alpha value, the higher the accuracy of the forecast values.
